---
title: "WAWBE_core_dataset_creation"
author: "Erin Dahl"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

print(Sys.time())


```

## Purpose

The goal of this code is to calculate wastewater trends and wastewater viral activity levels. 

## Libraries

```{r libraries, message=FALSE}

library(tidyverse)
library(REDCapR)
library(slider)
library(here)
library(sf)

time_start <- now()

```


## Load paths and data

```{r load_data}

# file containing redcap credentials, see credentials_ref.R for the variables needed
source(here("R/credentials_ref.R"))

# load the population df for the sewersheds, please contact our team for your file
load(here("population.RData"))

# load wval functions
source(here("R/wval_functions_v2.R"))

# source simple metric calculation functions
source(here("R/simple_metric_fns.R"))

# source calculation functions from nwss
source(here("R/redcap_samples_functions.R"))

# add county name here 
COUNTY <- "ENTER YOUR COUNTY NAME HERE"

```

## Read and merge redacap data

Read in redcap data and explore current redcap variables read in to R

```{r redcap}
# Step 1: Create wastewater sample dataset from REDCap data 

## Step 1a: Create dataset that contain wastewater sample data from REDCap
# retrieve the ww samples redcap data
redcap_ww_samples <- redcap_read(redcap_uri=redcap_url, token=ww_samples_token)$data

# ## Step 1b: Create subset wastewater sample dataset that filters out samples from repeating instruments 
# redcap_samples<- redcap_ww_samples %>% 
#                       filter(is.na(redcap_repeat_instrument)) %>%
#                       # remove columns
#                       select(-c(redcap_repeat_instrument,
#                                 test_result_date,
#                                 redcap_repeat_instance,
#                                 starts_with("pcr"),
#                                 lod_sewage,
#                                 quality_flag,
#                                 contains(c("variant", "complete", "timestamp", 
#                                            "received", "printed",
#                                            "orig")))) 
# 
# ## Step 1c. Create a subset of the wastewater samples dataset that coincides with WAWBE pcr targets and gene targets
# # pcr target associated data from c3_pcr_target_report_form
# redcap_targets <- redcap_ww_samples %>% 
#                       filter(redcap_repeat_instrument =="c3_pcr_target_report_form") %>%
#                       # keep only desired columns
#                       select(sample_id, 
#                              test_result_date,
#                              starts_with("pcr"),
#                              lod_sewage,
#                              quality_flag) %>%
#                       # map values
#                       mutate(pcr_target = case_match(pcr_target, 
#                                                       1 ~	"sars-cov-2",
#                                                       2	~ "delta",
#                                                       3	~ "omicron",
#                                                       4	~ "hMPXV",
#                                                       5	~ "FLUAV",
#                                                       6	~ "FLUBV",
#                                                       7 ~ "RSV"),
#                              pcr_gene_target = case_match(pcr_gene_target,
#                                                           1	~ "n1",
#                                                           2	~ "n2",
#                                                           3	~ "p681r",
#                                                           4	~ "del156-157",
#                                                           5	~ "wt214",
#                                                           6	~ "del69/70",
#                                                           7	~ "ins214epe",
#                                                           8	~ "del142-144",
#                                                           9	~ "wt156-157",
#                                                           10 ~ "e9l-nvar",
#                                                           11 ~ "G2R_G",
#                                                           12 ~ "InfA1",
#                                                           13 ~ "InfA2",
#                                                           14 ~ "InfB",
#                                                           15 ~ "n",
#                                                           16 ~ "InfA1 and InfA2 combined",
#                                                           17 ~ "RSV-A and RSV-B combined"))
# 
# 
# ## Step 1d: Create new dataset from merge of wastewater samples dataset and facility/site information dataset
# # merge the target specific and sample specific variables
# redcap_all_samples <- redcap_targets %>% 
#                           merge(redcap_samples, by = "sample_id") %>%
#                           mutate(sample_id = as.character(sample_id)) %>%
#                           # rename the sample site name to match naming in the population df
#                           rename(site_id = sample_site_name) %>% 
#                           # join to add the population data
#                           left_join(population, by = "site_id")  %>% 
#                           # format the date
#                           mutate(sample_collect_date = as.Date(sample_collect_date)) %>% 
#                           # remove data missing sample_site_ids
#                           filter(!is.na(site_id)) %>%
#                           # group by site id and gene target
#                           group_by(site_id,pcr_gene_target, .drop = F) %>% 
#                           # keep only first ww record if there are two for a site on the same day
#                           distinct(sample_collect_date, .keep_all = TRUE) %>%
#                           # targets tested, with no positive detection will
#                           # have a missing conc value. To show no detection, change NA to 0
#                           mutate(pcr_target_avg_conc = if_else(is.na(pcr_target_avg_conc),0, pcr_target_avg_conc))


## Step 1b: Transform redcap instrument data to df format

redcap_df <- redcap_samples_to_df(redcap_ww_samples)

## Step 1c:mapped and cleaned dataframe 
clean_results <- redcap_df_clean(redcap_df, population)

# Extract cleaned results from list
redcap_all_samples <- clean_results$redcap_all_samples_clean
seq_results <- clean_results$seq_results_clean

```


## Calculate wastewater trends

 - normalized flow
 - log normalized flow
 - determine trend categories

```{r ww_trends}
# Step 2: Create datasets that contain concentrations normalized by flow rate and population, calculate the 7-day rolling geometric mean of normalized adjusted concentrations, and calculate 15-day percent change of log-transformed normalized concentrations

geomean_pcr_gene_targets <- c("n", "InfA1 and InfA2 combined", "InfB", "RSV-A and RSV-B combined", "Norovirus GII")

# Step 2a: Create a new dataset that:
## - maps wastewater sample data to site IDs and population counts
## - creates new flow rate variable that converts flow rate from million gallons per day to liters per day
## - creates new adjusted concentration variable by applying limit of detection adjustment to sample concentrations that are less than the LOD sewage value of 0.35 gene copies/uL
## - creates new normalized concentration variable by multiplying by daily flow rate in liters and dividing that value by the sewersheds population
## - creates new log10-transformed normalized concentration variable for 15-day percent change calculation
## - filters dataset to log10-transformed normalized concentrations above zero. Note that this is because zeros will make the 15-day percent change function fail.
wwdata_flow_calc <- redcap_all_samples %>%  
                          # calculate flow rate in Liters
                          # original flow rate is in Million gallons per day; 
                          # 1 million gallon/day [US] to liters per day = 3785411.8 liters per day updated 2022-09-12
                          # If the pcr_target_avg_conc is less than the lod_sewage, 
                          # then the adjusted concentration value is 1/2 the lod sewage value
                          # This value change is is alignment with how CDC NWSS handles samples with concentrations below lod
                            mutate(flow_rate_l = flow_rate*3785411.8, 
                                   pcr_target_avg_conc_adj = if_else(pcr_target_avg_conc < lod_sewage, 0.5 * lod_sewage, pcr_target_avg_conc),
                                   pcr_target_norm_conc = (pcr_target_avg_conc_adj*flow_rate_l)/totalpop, 
                                   pcr_target_log_norm_conc = log10(pcr_target_norm_conc)) %>%  
                            # filter data with log normalized conc > 0
                            filter(pcr_target_log_norm_conc > 0) 

## Step 2b: Create dataset of all non-zero raw unadjusted concentrations to minimum normalized log-transformed concentrations by site and gene target and create a new adjustment value for non-detect replacement: 
# - this is because the 1/2*LOD adjustment across all samples with concentrations below LOD throws off 15-day % change metric with alternating detect/non-detects causing very high percent changes at some sites, especially given that this 1/2*LOD adjusted value is multiplied by flow rate when normalizing. 
# - replacing raw pcr target concentration values of 0 with 3/4 of the minimum log-10 transformed site/gene target specific normalized concentrations allows for the replaced concentration to be closer to 0 when flow rate is already taken into account (therefore, a high flow rate will not artificially increase the adjusted concentration)
# - this new log-10 transformed and adjusted normalized concentration is then fed into the 15-day percent change function
# - create new variable that is 3/4 of the minimum of normalized concentration for replacing zeros in 7-day rolling geometric mean
# Calculate the adjustment values for special handling of reported 0 concentration
adjustment_values <- wwdata_flow_calc %>% 
  filter(pcr_target_avg_conc > 0) %>% 
  group_by(pcr_gene_target, site_id) %>% 
  reframe(adj_norm_conc = 0.75 * min(pcr_target_norm_conc)) %>%
# fill in sites and gene target combos with zero for combos that have never had pcr_target_avg_conc > 0 
# these would have been filtered out above for low prevalence pathogens such as mpox, H5, H1 at some sites
# set the adjusted conc to 0
  complete(
    pcr_gene_target  = unique(wwdata_flow_calc$pcr_gene_target),
    site_id   = unique(wwdata_flow_calc$site_id),
    fill = list(adj_norm_conc = 0)
  ) %>% 
  ungroup()
## Step 2c: Create dataset that joins the newly created adjustment values dataset with the flow-pop normalized and log10 transformed wastewater sample dataset
# - when the raw pcr target concentration is 0, replace that with the adjusted normalized concentration values
# - raw concentrations above 0 but below LOD are still replaced with 1/2*LOD and then normalized and log-transformed
# - non-log transformed adjusted normalized concentration is used for 7-day rolling geometric mean
wwdata_flow_calc_adjusted <- wwdata_flow_calc %>%
  left_join(adjustment_values, by = c("pcr_gene_target", "site_id")) %>%
  mutate(
    pcr_target_norm_conc = if_else(pcr_target_avg_conc == 0, 
                                       adj_norm_conc, 
                                       pcr_target_norm_conc)) %>%
  select(-adj_norm_conc) # Remove the new_value_conc column after the update


## Step 2d: Create a new adjusted wastewater concentration dataset that:
# - calculates the 7-day rolling geometric mean of normalized concentrations for each site and gene target
# - have to remove missing sample dates otherwise the function will not work
# - gene target/sample dates with raw concentrations of zero are set to a geometric mean of zero
rolling_result <- wwdata_flow_calc_adjusted %>%
  filter(!is.na(sample_collect_date)) %>%
  # arrange by sample collect date
  arrange(sample_collect_date) %>%
  # group by site and gene target
  group_by(site_id, pcr_gene_target) %>%
  # calculate a 7-day rolling geometric mean of normalized concentrations with zeros replaced
  mutate(pcr_target_norm_7day_geomean = slide_index_dbl(
    .x = pcr_target_norm_conc,
    .i = sample_collect_date,
    .f = geo_mean,
    .before = days(6), # 6 previous days + current day = 7-day window
    .complete = TRUE
  )
) %>%
  mutate(pcr_target_norm_7day_geomean = if_else(pcr_target_avg_conc == 0,
                                                0,
                                                pcr_target_norm_7day_geomean)) %>%
  ungroup()

# Re-add the rows with missing sample dates to the dataset
# - these will have missing geometric means
wwdata_flow_calc_adjusted <- wwdata_flow_calc_adjusted %>%
    filter(is.na(sample_collect_date)) %>%
  mutate(pcr_target_norm_7day_geomean = NA_real_) %>%
  bind_rows(rolling_result)


```

## Calculate wastewater viral activity levels for site and county aggregations


```{r}
# Step 3: Create datasets that designate major lab methods and calculate wastewater viral activity levels at the site and county aggregation levels

wval_pcr_gene_targets <- c("n", "InfA1 and InfA2 combined", "RSV-A and RSV-B combined")

## Step 3a: Create dataset that adds the following variables to the wastewater sample dataset:
# - filters to micro lab name corresponding with PHL
# - adds pcr type variable that designates what type of pcr each instrument is performing
# - creates a variable that determines what the major lab method is based on the concentration method and the pcr type
wwdata_flow_calc_doh <- wwdata_flow_calc %>%
  rename(sample_site_id = site_id,
         sample_site_name = site) %>%
  # filter to PHL
  filter(micro_lab_name == '001') %>%
  # get pcr type from equipment ID
  mutate(pcr_type = case_when(equipment_id ==
    'QX1-KF-CeresNanotrap' ~ 'ddpcr',
    equipment_id ==
    'QIAcuityOne5-PlexDigitalPCRSystem' ~ 'qiagen dpcr',
    equipment_id ==
    'QIAcuityKFCeresNtrap' ~ 'qiagen dpcr'
  )) %>%
  # create major_lab_method variable
  mutate(major_lab_method = case_when(
    # condition for 'mlm' == 6
    concentration_method == 'skimmilk' ~ 6,
    # condition for 'mlm' == 7
    concentration_method == 'ceresnano' & pcr_type == 'ddpcr' ~ 7,
    # condition for 'mlm' == 8
    concentration_method == 'ceresnano' & pcr_type == 'qiagen dpcr' ~ 8,
    # default value if none match
    TRUE ~ NA_integer_
  )) 
  
## Step 3b: Create new site-week aggregated wastewater viral activity level for sars-cov-2 and flu a
# - these datasets will be further aggregated into county week level
# create overall wval dataset list for sars
wval_site_agg_raw_sars <- wval_v2(wwdata_flow_calc_doh, target = 'sars-cov-2')
wval_site_agg_raw_flua <- wval_v2(wwdata_flow_calc_doh, target = 'FLUAV')
wval_site_agg_raw_rsv <- wval_v2(wwdata_flow_calc_doh, target = 'RSV')

## Step 3c: Join both wval datasets together
# bind datasets
wval_site_agg_raw <- bind_rows(wval_site_agg_raw_sars, wval_site_agg_raw_flua, wval_site_agg_raw_rsv)

## Step 3d: Create clean site-week aggregated wval dataset and prep for merging
# select the columns to keep and rename for merging with core dataset
wval_site_agg <- wval_site_agg_raw %>% distinct() %>% 
  select(site_id, pcr_target, week_end, county_names, wval_site_wk, wval_site_wk_level) %>%
  # rename to match variables in wwdata 
  rename(county_main = county_names)

## Step 3e: Create new county-week aggregated wval dataset
# calculate county wval aggregations
wval_county_agg <- wval_site_agg_raw  %>%
  # calculate wval at the county-week aggregation
  wval_county_wk_agg() %>% 
  rename(county_main = county_names)

```


## Create finalized data object

```{r core_data}
# Step 4: Create final dataset

## Step 4a: Create clean wastewater sample dataset
keep_flow_calc <- wwdata_flow_calc_adjusted %>% select(sample_id,
                                                  site_id, 
                                                  site,
                                                  sample_collect_date,
                                                  test_result_date,
                                                  import_redcap_date,
                                                  micro_lab_name,
                                                  pcr_target,
                                                  pcr_gene_target,
                                                  pcr_target_avg_conc,
                                                  pcr_target_norm_conc,
                                                  pcr_target_norm_7day_geomean,
                                                  pcr_target_below_lod,
                                                  pcr_target_detection,
                                                  flow_rate_l,
                                                  totalpop,
                                                  quality_flag,
                                                  quality_flag_notes,
                                                  viral_data_flag,
                                                  data_flag_notes,
                                                  county_main) 

## Step 4b: Calculate the max sample date that has test results 
keep_flow_calc$sample_collect_date <- as.Date(keep_flow_calc$sample_collect_date)

## Step 4c: Create dataset that filters out any observations that do not have test results yet
# Get the max result date
max_result_date <- keep_flow_calc %>%
  filter(!is.na(test_result_date)) %>%
  pull(sample_collect_date) %>%
  max(na.rm = TRUE)

## Step 4d: Create a new dataset that:
# - creates a week_end variable for aggregating by week
# - joins wastewater trends dataset, wvals datasets, and wastewater sample datasets together
# - ensures that county and week wval datasets can align by week end, site, county, and pcr gene target
# - creates variable that designates when there is no recent data for calculating wvals
# - creates a new variable that designates when wval was not calculated (for influenza b, mpox, rsv, etc)
# - excludes mpox from 7-day rolling geometric mean calculation since there is not enough concentration data 
core_dataset <-  keep_flow_calc %>%
  mutate(week_end = sample_collect_date + days(7 - wday(sample_collect_date))) %>%
  # Adding in the gene target for the sake of merging
  mutate(
    pcr_target = case_when(
      pcr_gene_target %in% c('n', 'n1', 'n2') ~ "sars-cov-2",
      pcr_gene_target == "InfA1 and InfA2 combined" ~ 'FLUAV',
      pcr_gene_target == "RSV-A and RSV-B combined" ~ 'RSV',
      TRUE ~ pcr_target
    ),
    # Only display 7-day rolling geo mean for appropriate targets
    pcr_target_norm_7day_geomean = case_when(
      !(pcr_gene_target %in% geomean_pcr_gene_targets) ~ NA_real_,
      pcr_target_norm_7day_geomean == "NaN" ~ NA_real_,
      TRUE ~ pcr_target_norm_7day_geomean
    )
  ) %>%
  left_join(
    x = .,
    wval_site_agg,
    by = c("pcr_target", "week_end", "site_id", "county_main")
  ) %>%
  left_join(
    x = .,
    wval_county_agg,
    by = c("pcr_target", "week_end", "county_main")
  ) %>%
  mutate(pcr_target = if_else(
    pcr_gene_target %in% wval_pcr_gene_targets &
      is.na(sample_id),
    NA,
    pcr_target
  )) %>%
  # add "No Recent Data" or "Not Calculated" categorization to NAs
  mutate(
    wval_site_wk_level = case_when(
      pcr_gene_target %in% wval_pcr_gene_targets &
        is.na(wval_site_wk) &
        sample_collect_date <= max_result_date ~ "No Recent Data",
      !(pcr_gene_target %in% wval_pcr_gene_targets) &
        is.na(wval_site_wk) &
        sample_collect_date <= max_result_date ~ "Not Calculated",
      TRUE ~ wval_site_wk_level
    ),
    wval_county_wk_level = case_when(
      pcr_gene_target %in% wval_pcr_gene_targets &
        is.na(wval_county_wk) &
        sample_collect_date <= max_result_date ~ "No Recent Data",
      !(pcr_gene_target %in% wval_pcr_gene_targets) &
        is.na(wval_county_wk) &
        sample_collect_date <= max_result_date ~ "Not Calculated",
      TRUE ~ wval_county_wk_level
    )
  )


```

## Save core_dataset

```{r save}
# The line below saves an R object, alternatively the df could be saved as a csv
# save(core_dataset, file = core_dataset_path)

time_end <- now()

# total time to run script
time_end - time_start
```
